Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-08-02 threshold/draft.Rnw
restart in Alg. 2.1

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
Department of Mathematics,\\University of Rhode Island
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\}.
\]
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

We relax the column mean and scale assumptions without introducing much additional
computational or storage overhead as follows.  Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.

>>>>

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
Department of Mathematics,\\University of Rhode Island
\and
Michael Kane\\
Yale University
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\}.
\]
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
restart IRLBA to compute 
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows.  Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
